---
title: "Adjusting for Covariates"
author: "dfermin"
date: "3/25/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


This markdown describes how to adjust for known covariates in an omics data set.

```{r}
suppressMessages(library(dplyr))
suppressMessages(library(ggplot2))
suppressMessages(library(cancerdata))
```

## 1. Load example data set
To learn more about this data go to [http://www.bioconductor.org/packages/release/data/experiment/html/cancerdata.html](http://www.bioconductor.org/packages/release/data/experiment/html/cancerdata.html)

Each sample has 3 variable:

- age
- survival
- erp

This code loads the data and creates the necessary expression matrix and phenotype data.frame
```{r, error=F, warning=F, message=F}

data(VEER1)

pheno <- pData(VEER1)
j <- which(pheno$class == "NODM")
k <- which(pheno$class == "DM")
pheno$age <- NA
pheno$survival <- NA
pheno$erp <- NA

pheno$age[j] <- sapply(as.character(pheno$info[j]), function(x) { as.numeric(strsplit(x, split="\\.")[[1]][11]) })
pheno$survival[j] <- sapply(as.character(pheno$info[j]), function(x) { as.numeric(strsplit(x, split="\\.")[[1]][7]) })
pheno$erp[j] <- sapply(as.character(pheno$info[j]), function(x) { strsplit(x, split="\\.")[[1]][14] })

pheno$age[k] <- sapply(as.character(pheno$info[k]), function(x) { as.numeric(strsplit(x, split="\\.")[[1]][12]) })
pheno$survival[k] <- sapply(as.character(pheno$info[k]), function(x) { as.numeric(strsplit(x, split="\\.")[[1]][8]) })
pheno$erp[k] <- sapply(as.character(pheno$info[k]), function(x) { strsplit(x, split="\\.")[[1]][15] })

pheno$erp <- as.factor(pheno$erp)
pheno$class <- as.factor(pheno$class)

pheno <- select(pheno, -info) ## don't need this anymore
rownames(pheno) <- paste0("Sample", seq(nrow(pheno)))

mat <- exprs(VEER1)
mat <- mat[grep("NM_", rownames(mat)),]
colnames(mat) <- paste0("Sample", seq(ncol(mat)))

## remove rows of 'mat' that contain NaN
rs <- apply(mat, 1, sum)
j <- which(is.nan(rs))
mat <- mat[-j,]


print(head(pheno))

print(mat[1:5,1:8])
```

## 2. Adjust for 3 variables

We will use linear regression to adjust for the 3 variables mentioned above.
We use the observed expression data in `mat` as the **outcome** variable in the model.
You have to do this for each sample in the matrix.

If the expression matrix is really large you can speed this up using the `doParallel` package.
```{r}

out <- mat
out[, 1:ncol(out) ] <- NA

nr <- nrow(mat)
for(j in 1:nr) {
  
  ## create a data.frame where each row is one patient
  ## The Y variable is the observed expression data for the current gene
  ## at row 'j' of 'mat'
  tmpX <- data.frame(Y=mat[j,], dplyr::select(pheno, -class))
  
  ## run linear regression with the expression value as the response variable
  tmp.fit <- lm(Y ~ ., data=tmpX)

  ## the residuals of the linear model 'tmp.fit' are the data adjusted for the 
  ## covariates we are concerned about. In this case 'age', 'erp' and 'survival'
  out[j,] <- round(residuals(tmp.fit), 4)
}


## 'out' has the expression data after adjusting for age, erp and survival

print(out[1:5,1:8])
```

